---
title: "Measles outbreak hubs (2024–2026) vs historical persistence and MMR coverage"
output: html_document
---

# Setup
```{r}
library(tidyverse)   # Data wrangling + plotting (dplyr, tidyr, ggplot2, readr, stringr)
library(readxl) #allows reading of excel files
library(lubridate)   # Date parsing and extracting years
library(janitor)     # clean_names() for consistent column names
library(scales)      # percent() formatting
library(here)        # Stable project-root paths
```


# Loading cdc-measlesbystate.csv
```{r}
#Load in raw CDC measles data
cdcmeaslesbystate_raw <- read_csv(here("data/raw-data/cdc-measlesbystate.csv"), show_col_types = FALSE) |>
  clean_names()

#View the first 8 rows in the raw data
names(cdcmeaslesbystate_raw)
cdcmeaslesbystate_raw |> slice_head(n = 8)
```
# Cleaning
```{r}
#Rename the location column to state
  if ("location" %in% names(cdcmeaslesbystate_raw)) {
  cdcmeaslesbystate_raw <- cdcmeaslesbystate_raw |> rename(state = location)
}

#Rename columns while converting the dataset from a wide format to a long format
  cdcmeaslesbystate_long <- cdcmeaslesbystate_raw |>
  select(state, matches("^x?\\d{4}$")) |>
  pivot_longer(
    cols = matches("^x?\\d{4}$"),
    names_to = "year_raw",
    values_to = "cases"
  ) |>
  mutate(
    year  = readr::parse_number(year_raw), #extract year numbers from string values
    cases = suppressWarnings(as.numeric(cases)), #convert case variable to numeric
    state = stringr::str_to_title(stringr::str_squish(state)) #Format states to be capitalized
  ) |>
  filter(
    !is.na(state), 
    !is.na(year), 
    state != "District Of Columbia") #Remove rows associated with District of Columbia

# Quick checks
names(cdcmeaslesbystate_long)
cdcmeaslesbystate_long |> count(year)
cdcmeaslesbystate_long |>
  filter(state %in% c("Texas", "South Carolina", "Arizona")) |>
  arrange(state, year)
```
# Summary of cleaned cdc-measlesbystate.csv
```{r}
recent_state <- cdcmeaslesbystate_long |>
  group_by(state) |>
  summarise(
    total_cases_2024_26 = sum(cases, na.rm = TRUE),
    peak_cases_2024_26  = max(cases, na.rm = TRUE),
    mean_cases_2024_26  = mean(cases, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_cases_2024_26))

recent_state |> slice_head(n = 15)
summary(recent_state$total_cases_2024_26)
```

# Loading measles-noncumulative.csv
```{r}
#Load in historical measles data
hist_raw <- read_csv(here("data/raw-data/measles-noncumulative.csv"), show_col_types = FALSE) |>
  clean_names()

#View the names of the first 3 columns in the raw historical data
names(hist_raw)
hist_raw |> slice_head(n = 3)
```
# Cleaning: Weekly totals to yearly totals per state
```{r}
#Clean the raw data and make a new dataframe containing yearly measles data
hist_yearly <- hist_raw |>
  transmute( #create new dataframe containing these variables
    state = str_to_title(str_squish(admin1name)), #convert states to be in similar format to previous dataset
    start_date = ymd(period_start_date), #convert period state date variable to a proper start date
    year = year(start_date), #extract year from start_date
    cases = as.numeric(count_value) #make count_value a numeric variable
  ) |>
  filter(
    !is.na(state), 
    !is.na(year), 
    !is.na(cases),
    !state %in% c("American Samoa", "Guam", "District of Columbia", "Northern Mariana Islands", "Puerto Rico", "Virgin Islands, U.s.")) |>
    #remove cities and U.S. territories
  group_by(state, year) |> #group data by state and year
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") #sum up number of cases by year for each state

#Check to confirm data is cleaned
hist_yearly |> summarise(min_year = min(year), max_year = max(year), states = n_distinct(state))
hist_yearly |> slice_head(n = 10)
```
# Narrow down to the years 1980-2009 to look at how long cases persisted in each state
```{r}
start_year <- 1980
end_year   <- 2009
```
# Function to answer the question "What is the longest number of consecutive years a state had measles cases"?
```{r}
longest_run_nonzero_hist <- function(x) {
  r <- rle(x)
  if (!any(r$values)) return(0L)
  max(r$lengths[r$values])
}
```

```{r}
# Create a state-level historical measles summary table (one row per state)
hist_state <- hist_yearly |>
# Keep only the years we want to study (1980–2009)
  filter(year >= start_year, year <= end_year) |>
# Ensure EVERY state has EVERY year in the range (fills missing state-years)
# If a state-year is missing, we assume 0 cases for that year
  complete(state, year = seq(start_year, end_year), fill = list(cases = 0)) |>
# Do the next calculations separately for each state
  group_by(state) |>
# Sort years in order so streak calculations (runs) work correctly
  arrange(year) |>
# Collapse each state's many year-rows into ONE summary row of features   
  summarise(
    # Share of years in the window where the state had ANY measles cases
    # cases > 0 becomes TRUE/FALSE, and mean(TRUE/FALSE) becomes a proportion
    pct_years_any = mean(cases > 0),
# Average number of measles cases per year across the time window  
    mean_cases_80_09 = mean(cases),
 # Largest single-year measles case count across the time window
 # (Uses the helper function you defined earlier)
    max_cases_80_09  = max(cases),
    longest_run_nonzero = longest_run_nonzero_hist(cases > 0),
# Drop grouping so hist_state is a normal (ungrouped) dataframe
    .groups = "drop"
  )
# Show the 12 states with the highest share of years that had measles cases
hist_state |> arrange(desc(pct_years_any)) |> slice_head(n = 12)
# Show the 12 states with the longest continuous streak of years with measles cases
hist_state |> arrange(desc(longest_run_nonzero)) |> slice_head(n = 12)
```


# Loading and Cleaning MMR data
```{r}
# load the single MMR file (contains all years) 
mmr_raw <- read_csv(here("data/raw-data/cdc-mmr2024.csv"), show_col_types = FALSE) |>
  clean_names()

# Quick look at structure
names(mmr_raw)
mmr_raw |> slice_head(n = 10)

# Clean core variables
mmr_clean <- mmr_raw |>
  transmute(
    state = str_to_title(str_squish(location)),                 # standardize state names
    school_year = str_squish(school_year),                      # trim spaces
    school_year_start = as.integer(str_extract(school_year, "\\d{4}")), # extract starting year
    mmr_pct = readr::parse_number(estimated_percent_vaccinated),# handles "95.4%" or "95.4"
    category = if ("category" %in% names(mmr_raw)) category else NA_character_
  ) |>
  filter(!is.na(state), 
  !is.na(school_year_start), 
  !is.na(mmr_pct),
  !state %in% c(
    "Houston", "San Antonio", "New York City", "District Of Columbia" 
  )
  )

# Check which years exist
mmr_clean |> count(school_year, school_year_start) |> arrange(school_year_start)

# Collapse to ONE row per state-year (handles duplicate categories/rows)
mmr_state_year <- mmr_clean |>
  group_by(state, school_year_start) |>
  summarise(
    mmr_pct = mean(mmr_pct, na.rm = TRUE),   # average across duplicates if present
    n_rows = n(),                             # how many rows contributed
    .groups = "drop"
  )

# Confirm we truly have one row per state-year now
mmr_state_year |> count(school_year_start) |> arrange(school_year_start)
mmr_state_year |> filter(state %in% c("Texas","South Carolina")) |> arrange(state, school_year_start)
```

# Loading of State Educational Attainment Data
```{r}
#Load in data using read_excel()
education_raw <- read_excel(here("data/raw-data/educationattainment.xlsx"), skip = 6) |>
  clean_names()

#Remove NA columns due to how the data is loaded in
education_clean <- education_raw %>% select(-c("x3", "x6", "x9", "x12", "x15", "x18", "x21", "x24", "x27", "x30", "x33", "x36", "x39", "x42"))

#Rename the columns due to the structure of the excel datasheet
education_clean <- rename(
  education_clean,
  "state" = "x1",
  "2012_18to24hs_percent" = "x2",
  "2012_18to24hs_sd" = "x4",
  "2022_18to24hs_percent" = "x3_2",
  "2022_18to24hs_sd" = "x7",
  "2012_25pluslessthanhs_percent" = "x4_2",
  "2012_25pluslessthanhs_sd" = "x10",
  "2012_25plushsorhigher_percent" = "x5",
  "2012_25plushsorhigher_sd" = "x13",
  "2012_25plustotalbachgrad_percent" = "x6_2",
  "2012_25plustotalbachgrad_sd" = "x16",
  "2012_25plusbachonly_percent" = "x7_2",
  "2012_25plusto24bachonly_sd" = "x19",
  "2012_25plusgraddegree_percent" = "x8",
  "2012_25plusto24graddegree_sd" = "x22",
  "2022_25pluslessthanhs_percent" = "x9_2",
  "2022_25pluslessthanhs_sd" = "x25",
  "2022_25plushsorhighertotal_percent" = "x10_2",
  "2022_25plushsorhighertotal_sd" = "x28",
  "2022_25plushsorhigherhsonly_percent" = "x11",
  "2022_25plushsorhigherhsonly_sd" = "x31",
  "2022_25plussomecolldegree_percent" = "x12_32",
  "2022_25plussomecolldegree_sd" = "x12_34",
  "2022_25plusbachorhighertotal_percent" = "x13_2",
  "2022_25plusto24bachorhighertotal_sd" = "x37",
  "2022_25plusbachonly_percent" = "x14",
  "2022_25plusto24bachonly_sd" = "x40",
  "2022_25plusgraddegree_percent" = "x15_2",
  "2022_25plusto24graddegree_sd" = "x43"
)

#Remove rows of data not affiliated with the 50 U.S. States
##Make a df which contains United States, District of Columbia, Other Jurisdictions, Puerto Rico and final three rows
rows_to_remove <- c(1, 10, 53, 54, 55, 56, 57)

##Use rows_to_remove to remove the certain jurisdictions using slice()
education_clean <- education_clean %>% slice(-rows_to_remove)

#Add a column which calculates the difference between 2012 and 2022 high school graduation rates for individuals aged 18-24
education_clean <- education_clean %>%
  mutate(`hsgradratediff` = `2022_18to24hs_percent` - `2012_18to24hs_percent`
  )
```

# Exploratory summaries and plots from cleaned data
```{r}
#Exploratory summaries for MMR data

#Find states with the lowest recent MMR coverage (chose last available year)
last_year <- max(mmr_state_year$school_year_start, na.rm = TRUE)

mmr_state_year |>
  filter(school_year_start == last_year) |>
  arrange(mmr_pct) |>
  select(state, school_year_start, mmr_pct) |>
  slice_head(n = 15)

#Determine states that were below 95% most often (across all years)
mmr_state_features <- mmr_state_year |>
  group_by(state) |>
  summarise(
    mmr_mean = mean(mmr_pct, na.rm = TRUE),
    mmr_min  = min(mmr_pct, na.rm = TRUE),
    frac_years_below_95 = mean(mmr_pct < 95, na.rm = TRUE),
    first_year = min(school_year_start),
    last_year  = max(school_year_start),
    n_years = n(),
    .groups = "drop"
  ) |>
  arrange(desc(frac_years_below_95))

mmr_state_features |> slice_head(n = 20)

#Calculate national mean and median over time
mmr_national_trend <- mmr_state_year |>
  group_by(school_year_start) |>
  summarise(
    median_mmr = median(mmr_pct, na.rm = TRUE),
    mean_mmr = mean(mmr_pct, na.rm = TRUE),
    .groups = "drop"
  )

mmr_national_trend

#Summarize educational attainment across 50 U.S. states
summary(education_clean)
```
# Joining all four cleaned data into a master table
```{r}
features <- recent_state |>

  # Join historical persistence features
  left_join(hist_state, by = "state") |>

  # Join education attainment features
  left_join(education_clean, by = "state") |>

  # Join vaccination features
  left_join(mmr_state_features, by = "state")
  features |> glimpse()
  features |> summarise(across(where(is.numeric), ~mean(is.na(.))))
```
```{r}
# Exploratory plots

# Plot 1a: National trend (median MMR coverage across states)
mmr_national_trend |>
  ggplot(aes(x = school_year_start, y = median_mmr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Median kindergarten MMR coverage Nationally",
    x = "School year (start)",
    y = "Median MMR coverage (%)",
    caption = "Dashed line = 95% (herd-immunity target for measles)"
  ) +
  theme_minimal()

# plot 1b:  National trend (median MMR coverage across states)
mmr_national_trend |>
  ggplot(aes(x = school_year_start, y = mean_mmr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Mean kindergarten MMR coverage Nationally",
    x = "School year (start)",
    y = "Median MMR coverage (%)",
    caption = "Dashed line = 95% (herd-immunity target for measles)"
  ) +
  theme_minimal()
# Plot 2: MMR coverage overtime for specific states (edit list as needed)
# Y intercept = For measles, epidemiologists often cite ~95% vaccination coverage as the level needed for strong herd immunity.(STILL NEED TO ADD REF)
focus_states <- c("Texas","South Carolina","California","New York","Florida")

mmr_state_year |>
  filter(state %in% focus_states) |>
  ggplot(aes(
    x = school_year_start,
    y = mmr_pct,
    color = state,      
    group = state
  )) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Kindergarten MMR coverage over time (selected states)",
    x = "School year (start)",
    y = "MMR coverage (%)",
    color = "State"     # legend title
  ) +
  theme_minimal()
```

```{r}
# Exploratory summaries from recent measles data

# Exploratory plots
# Plot 1
recent_state |>
  arrange(desc(total_cases_2024_26)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, total_cases_2024_26),
             y = total_cases_2024_26)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top states by measles cases (2024–2026)",
    x = "State",
    y = "Total cases"
  ) +
  theme_minimal()
```

```{r}
# Plot 2
##Select certain states to hone in on
focus_states <- c("Texas","South Carolina","Arizona","Utah","California","New York")

#Plot the measles trends for the six selected states, some of which are having active measles outbreaks in the U.S.
cdcmeaslesbystate_long |>
  filter(state %in% focus_states) |>
  ggplot(aes(year, cases, color = state)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(min(cdcmeaslesbystate_long$year),
                 max(cdcmeaslesbystate_long$year),
                 by = 1)
  ) +
  labs(
    title = "Year-by-year recent trends for key states",
    x = "Year",
    y = "Cases"
  ) +
  theme_minimal()
```
```{r}
# Exploratory summaries from historic measles data

# Plot 1
hist_state |>
  arrange(desc(pct_years_any)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, pct_years_any),
             y = pct_years_any)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "States with most persistent measles presence (1980–2009)",
    x = "State",
    y = "Percent of years with cases"
  ) +
  theme_minimal()

# Plot 2
hist_state |>
  arrange(desc(longest_run_nonzero)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, longest_run_nonzero),
             y = longest_run_nonzero)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Longest consecutive measles streak by state",
    x = "State",
    y = "Years in a row with measles"
  ) +
  theme_minimal()
```

```{r}
# Exploratory summaries from educational attainment dataset

# Plot 1: 18 to 24 year old percent graduation from high school by state in 2012
education_clean |>
  arrange(desc(`2012_18to24hs_percent`)) |>
  ggplot(aes(x = `2012_18to24hs_percent`,
             y = reorder(state, `2012_18to24hs_percent`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Top states by high school graduation rates in 2012",
    x = "State",
    y = "Percent"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Plot 2: 18 to 24 year old percent graduation from high school by state in 2022
education_clean |>
  arrange(desc(`2022_18to24hs_percent`)) |>
  ggplot(aes(x = `2022_18to24hs_percent`,
             y = reorder(state, `2022_18to24hs_percent`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Top states by high school graduation rates in 2022",
    x = "State",
    y = "Percent"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#Plot 3: High school graduation rate differences over time
education_clean |>
  arrange(desc(`hsgradratediff`)) |>
  ggplot(aes(x = `hsgradratediff`,
             y = reorder(state, `hsgradratediff`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Graduation Rate Differences from 2012 to 2022",
    x = "State",
    y = "Percent Difference"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Exploratory data analysis

```{r}
# Recent measles cases vs average MMR coverage
# plot without linear regression
library(ggrepel)

features |>
  ggplot(aes(x = mmr_mean, y = total_cases_2024_26)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  scale_y_sqrt() +
  labs(
    title = "Recent measles cases vs average MMR coverage",
    x = "Average MMR coverage (%)",
    y = "Total measles cases 2024–2026 (sqrt scale)"
  ) +
  theme_minimal()
```

```{r}
# Recent measles cases vs average MMR coverage
# plot with linear regression
features |>
  ggplot(aes(x = mmr_mean, y = total_cases_2024_26)) +
  geom_point(size = 2) +
  
  # Add regression line with confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # Label each state without overlapping
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scale helps when some states have very high counts
  scale_y_sqrt() +
  
  labs(
    title = "Recent measles cases vs average MMR coverage",
    x = "Average MMR coverage (%)",
    y = "Total measles cases 2024–2026 (sqrt scale)"
  ) +
  theme_minimal()
```


```{r}
# Recent peak measles cases vs historical peak measles burden
# plot with linear regression
features |>
  ggplot(aes(x = max_cases_80_09, y = peak_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent peak measles cases vs historical peak measles burden",
    x = "Maximum yearly measles cases (1980–2009)",
    y = "Peak measles cases (2024–2026)"
  ) +
  
  theme_minimal()
```

```{r}
# MMR Vaccination Rates vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25pluslessthanhs_percent`, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "MMR Vaccine Rates correlated to 2022 rates of high school incompletion by state",
    x = "Incomplete High School Graduation Rates (%)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# Vaccination rates vs positive/negative difference in high school graduation rate by state
# plot with linear regression
features |>
  ggplot(aes(x = hsgradratediff, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent MMR vaccine rates correlated to differences in graduation rate between 2012 and 2022",
    x = "High School Graduation Rate Difference (%)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# MMR vaccination rate vs grad school completion
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25plusbachorhighertotal_percent`, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "MMR Vaccination Rates correlated to attaining higher education degree(s)",
    x = "Bacherlors or Higher Education Rate (percent)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# Peak measles case burden (2024-26) vs grad school completion
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25plusbachorhighertotal_percent`, y = peak_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent peak measles cases correlated to differences in graduation rate between 2012 and 2022",
    x = "High School Graduation Rate Difference (percent)",
    y = "Peak measles cases (2024–2026)"
  ) +

  theme_minimal()

# Total Measles Cases vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25pluslessthanhs_percent`, y = total_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Total Measles Cases (2024-26) correlated to 2022 rates of high school incompletion by state",
    x = "Incomplete High School Graduation Rates (%)",
    y = "Total Measles Cases (2024-26)"
  ) +

  theme_minimal()

# Total Measles Cases vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2012_18to24hs_percent`, y = total_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Total Measles Cases (2024-26) correlated to 2022 rates of high school incompletion by state",
    x = "2012 High School Graduation Rates for 18-24 year olds (%)",
    y = "Total Measles Cases (2024-26)"
  ) +

  theme_minimal()
```