---
title: "Measles outbreak hubs (2024–2026) vs historical persistence and MMR coverage"
output: html_document
---

# Setup
```{r}
library(tidyverse)   # Data wrangling + plotting (dplyr, tidyr, ggplot2, readr, stringr)
library(readxl) #allows reading of excel files
library(lubridate)   # Date parsing and extracting years
library(janitor)     # clean_names() for consistent column names
library(scales)      # percent() formatting
library(here)        # Stable project-root paths
```


# Loading cdc-measlesbystate.csv
```{r}
#Load in raw CDC measles data
cdcmeaslesbystate_raw <- read_csv(here("data/raw-data/cdc-measlesbystate.csv"), show_col_types = FALSE) |>
  clean_names()

#View the first 8 rows in the raw data
names(cdcmeaslesbystate_raw)
cdcmeaslesbystate_raw |> slice_head(n = 8)
```
# Cleaning
```{r}
#Rename the location column to state
  if ("location" %in% names(cdcmeaslesbystate_raw)) {
  cdcmeaslesbystate_raw <- cdcmeaslesbystate_raw |> rename(state = location)
}

#Rename columns while converting the dataset from a wide format to a long format
  cdcmeaslesbystate_long <- cdcmeaslesbystate_raw |>
  select(state, matches("^x?\\d{4}$")) |>
  pivot_longer(
    cols = matches("^x?\\d{4}$"),
    names_to = "year_raw",
    values_to = "cases"
  ) |>
  mutate(
    year  = readr::parse_number(year_raw), #extract year numbers from string values
    cases = suppressWarnings(as.numeric(cases)), #convert case variable to numeric
    state = stringr::str_to_title(stringr::str_squish(state)) #Format states to be capitalized
  ) |>
  filter(
    !is.na(state), 
    !is.na(year), 
    state != "District Of Columbia") #Remove rows associated with District of Columbia

# Quick checks
names(cdcmeaslesbystate_long)
cdcmeaslesbystate_long |> count(year)
cdcmeaslesbystate_long |>
  filter(state %in% c("Texas", "South Carolina", "Arizona")) |>
  arrange(state, year)
```
# Summary of cleaned cdc-measlesbystate.csv
```{r}
recent_state <- cdcmeaslesbystate_long |>
  group_by(state) |>
  summarise(
    total_cases_2024_26 = sum(cases, na.rm = TRUE),
    peak_cases_2024_26  = max(cases, na.rm = TRUE),
    mean_cases_2024_26  = mean(cases, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_cases_2024_26))

recent_state |> slice_head(n = 15)
summary(recent_state$total_cases_2024_26)
```

# Loading measles-noncumulative.csv
```{r}
#Load in historical measles data
hist_raw <- read_csv(here("data/raw-data/measles-noncumulative.csv"), show_col_types = FALSE) |>
  clean_names()

#View the names of the first 3 columns in the raw historical data
names(hist_raw)
hist_raw |> slice_head(n = 3)
```
# Cleaning: Weekly totals to yearly totals per state
```{r}
#Clean the raw data and make a new dataframe containing yearly measles data
hist_yearly <- hist_raw |>
  transmute( #create new dataframe containing these variables
    state = str_to_title(str_squish(admin1name)), #convert states to be in similar format to previous dataset
    start_date = ymd(period_start_date), #convert period state date variable to a proper start date
    year = year(start_date), #extract year from start_date
    cases = as.numeric(count_value) #make count_value a numeric variable
  ) |>
  filter(
    !is.na(state), 
    !is.na(year), 
    !is.na(cases),
    !state %in% c("American Samoa", "Guam", "District of Columbia", "Northern Mariana Islands", "Puerto Rico", "Virgin Islands, U.s.")) |>
    #remove cities and U.S. territories
  group_by(state, year) |> #group data by state and year
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") #sum up number of cases by year for each state

#Check to confirm data is cleaned
hist_yearly |> summarise(min_year = min(year), max_year = max(year), states = n_distinct(state))
hist_yearly |> slice_head(n = 10)
```
# Narrow down to the years 1980-2009 to look at how long cases persisted in each state
```{r}
start_year <- 1980
end_year   <- 2009
```
# Function to answer the question "What is the longest number of consecutive years a state had measles cases"?
```{r}
longest_run_nonzero_hist <- function(x) {
  r <- rle(x)
  if (!any(r$values)) return(0L)
  max(r$lengths[r$values])
}
```

```{r}
# Create a state-level historical measles summary table (one row per state)
hist_state <- hist_yearly |>
# Keep only the years we want to study (1980–2009)
  filter(year >= start_year, year <= end_year) |>
# Ensure EVERY state has EVERY year in the range (fills missing state-years)
# If a state-year is missing, we assume 0 cases for that year
  complete(state, year = seq(start_year, end_year), fill = list(cases = 0)) |>
# Do the next calculations separately for each state
  group_by(state) |>
# Sort years in order so streak calculations (runs) work correctly
  arrange(year) |>
# Collapse each state's many year-rows into ONE summary row of features   
  summarise(
    # Share of years in the window where the state had ANY measles cases
    # cases > 0 becomes TRUE/FALSE, and mean(TRUE/FALSE) becomes a proportion
    pct_years_any = mean(cases > 0),
# Average number of measles cases per year across the time window  
    mean_cases_80_09 = mean(cases),
 # Largest single-year measles case count across the time window
 # (Uses the helper function you defined earlier)
    max_cases_80_09  = max(cases),
    longest_run_nonzero = longest_run_nonzero_hist(cases > 0),
# Drop grouping so hist_state is a normal (ungrouped) dataframe
    .groups = "drop"
  )
# Show the 12 states with the highest share of years that had measles cases
hist_state |> arrange(desc(pct_years_any)) |> slice_head(n = 12)
# Show the 12 states with the longest continuous streak of years with measles cases
hist_state |> arrange(desc(longest_run_nonzero)) |> slice_head(n = 12)
```


# Loading and Cleaning MMR data
```{r}
# load the single MMR file (contains all years) 
mmr_raw <- read_csv(here("data/raw-data/cdc-mmr2024.csv"), show_col_types = FALSE) |>
  clean_names()

# Quick look at structure
names(mmr_raw)
mmr_raw |> slice_head(n = 10)

# Clean core variables
mmr_clean <- mmr_raw |>
  transmute(
    state = str_to_title(str_squish(location)),                 # standardize state names
    school_year = str_squish(school_year),                      # trim spaces
    school_year_start = as.integer(str_extract(school_year, "\\d{4}")), # extract starting year
    mmr_pct = readr::parse_number(estimated_percent_vaccinated),# handles "95.4%" or "95.4"
    category = if ("category" %in% names(mmr_raw)) category else NA_character_
  ) |>
  filter(!is.na(state), 
  !is.na(school_year_start), 
  !is.na(mmr_pct),
  !state %in% c(
    "Houston", "San Antonio", "New York City", "District Of Columbia" 
  )
  )

# Check which years exist
mmr_clean |> count(school_year, school_year_start) |> arrange(school_year_start)

# Collapse to ONE row per state-year (handles duplicate categories/rows)
mmr_state_year <- mmr_clean |>
  group_by(state, school_year_start) |>
  summarise(
    mmr_pct = mean(mmr_pct, na.rm = TRUE),   # average across duplicates if present
    n_rows = n(),                             # how many rows contributed
    .groups = "drop"
  )

# Confirm we truly have one row per state-year now
mmr_state_year |> count(school_year_start) |> arrange(school_year_start)
mmr_state_year |> filter(state %in% c("Texas","South Carolina")) |> arrange(state, school_year_start)
```

# Loading of State Educational Attainment Data
```{r}
#Load in data using read_excel()
education_raw <- read_excel(here("data/raw-data/educationattainment.xlsx"), skip = 6) |>
  clean_names()

#Remove NA columns due to how the data is loaded in
education_clean <- education_raw %>% select(-c("x3", "x6", "x9", "x12", "x15", "x18", "x21", "x24", "x27", "x30", "x33", "x36", "x39", "x42"))

#Rename the columns due to the structure of the excel datasheet
education_clean <- rename(
  education_clean,
  "state" = "x1",
  "2012_18to24hs_percent" = "x2",
  "2012_18to24hs_sd" = "x4",
  "2022_18to24hs_percent" = "x3_2",
  "2022_18to24hs_sd" = "x7",
  "2012_25pluslessthanhs_percent" = "x4_2",
  "2012_25pluslessthanhs_sd" = "x10",
  "2012_25plushsorhigher_percent" = "x5",
  "2012_25plushsorhigher_sd" = "x13",
  "2012_25plustotalbachgrad_percent" = "x6_2",
  "2012_25plustotalbachgrad_sd" = "x16",
  "2012_25plusbachonly_percent" = "x7_2",
  "2012_25plusto24bachonly_sd" = "x19",
  "2012_25plusgraddegree_percent" = "x8",
  "2012_25plusto24graddegree_sd" = "x22",
  "2022_25pluslessthanhs_percent" = "x9_2",
  "2022_25pluslessthanhs_sd" = "x25",
  "2022_25plushsorhighertotal_percent" = "x10_2",
  "2022_25plushsorhighertotal_sd" = "x28",
  "2022_25plushsorhigherhsonly_percent" = "x11",
  "2022_25plushsorhigherhsonly_sd" = "x31",
  "2022_25plussomecolldegree_percent" = "x12_32",
  "2022_25plussomecolldegree_sd" = "x12_34",
  "2022_25plusbachorhighertotal_percent" = "x13_2",
  "2022_25plusto24bachorhighertotal_sd" = "x37",
  "2022_25plusbachonly_percent" = "x14",
  "2022_25plusto24bachonly_sd" = "x40",
  "2022_25plusgraddegree_percent" = "x15_2",
  "2022_25plusto24graddegree_sd" = "x43"
)

#Remove rows of data not affiliated with the 50 U.S. States
##Make a df which contains United States, District of Columbia, Other Jurisdictions, Puerto Rico and final three rows
rows_to_remove <- c(1, 10, 53, 54, 55, 56, 57)

##Use rows_to_remove to remove the certain jurisdictions using slice()
education_clean <- education_clean %>% slice(-rows_to_remove)

#Add a column which calculates the difference between 2012 and 2022 high school graduation rates for individuals aged 18-24
education_clean <- education_clean %>%
  mutate(`hsgradratediff` = `2022_18to24hs_percent` - `2012_18to24hs_percent`
  )
```

# Exploratory summaries and plots from cleaned data
```{r}
#Exploratory summaries for MMR data

#Find states with the lowest recent MMR coverage (chose last available year)
last_year <- max(mmr_state_year$school_year_start, na.rm = TRUE)

mmr_state_year |>
  filter(school_year_start == last_year) |>
  arrange(mmr_pct) |>
  select(state, school_year_start, mmr_pct) |>
  slice_head(n = 15)

#Determine states that were below 95% most often (across all years)
mmr_state_features <- mmr_state_year |>
  group_by(state) |>
  summarise(
    mmr_mean = mean(mmr_pct, na.rm = TRUE),
    mmr_min  = min(mmr_pct, na.rm = TRUE),
    frac_years_below_95 = mean(mmr_pct < 95, na.rm = TRUE),
    first_year = min(school_year_start),
    last_year  = max(school_year_start),
    n_years = n(),
    .groups = "drop"
  ) |>
  arrange(desc(frac_years_below_95))

mmr_state_features |> slice_head(n = 20)

#Calculate national mean and median over time
mmr_national_trend <- mmr_state_year |>
  group_by(school_year_start) |>
  summarise(
    median_mmr = median(mmr_pct, na.rm = TRUE),
    mean_mmr = mean(mmr_pct, na.rm = TRUE),
    .groups = "drop"
  )

mmr_national_trend

#Summarize educational attainment across 50 U.S. states
summary(education_clean)
```
# Joining all four cleaned data into a master table
```{r}
features <- recent_state |>

  # Join historical persistence features
  left_join(hist_state, by = "state") |>

  # Join education attainment features
  left_join(education_clean, by = "state") |>

  # Join vaccination features
  left_join(mmr_state_features, by = "state")
  features |> glimpse()
  features |> summarise(across(where(is.numeric), ~mean(is.na(.))))

#Save Cleaned State Data for All Variables
saveRDS(features, "data/processed-data/measlesstatefeatures_data.rds")
```
```{r}
# Exploratory plots

# Plot 1a: National trend (median MMR coverage across states)
mmr_national_trend |>
  ggplot(aes(x = school_year_start, y = median_mmr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Median kindergarten MMR coverage Nationally",
    x = "School year (start)",
    y = "Median MMR coverage (%)",
    caption = "Dashed line = 95% (herd-immunity target for measles)"
  ) +
  theme_minimal()

# plot 1b:  National trend (median MMR coverage across states)
mmr_national_trend |>
  ggplot(aes(x = school_year_start, y = mean_mmr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Mean kindergarten MMR coverage Nationally",
    x = "School year (start)",
    y = "Median MMR coverage (%)",
    caption = "Dashed line = 95% (herd-immunity target for measles)"
  ) +
  theme_minimal()
# Plot 2: MMR coverage overtime for specific states (edit list as needed)
# Y intercept = For measles, epidemiologists often cite ~95% vaccination coverage as the level needed for strong herd immunity.(STILL NEED TO ADD REF)
focus_states <- c("Texas","South Carolina","California","New York","Florida")

mmr_state_year |>
  filter(state %in% focus_states) |>
  ggplot(aes(
    x = school_year_start,
    y = mmr_pct,
    color = state,      
    group = state
  )) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(
    title = "Kindergarten MMR coverage over time (selected states)",
    x = "School year (start)",
    y = "MMR coverage (%)",
    color = "State"     # legend title
  ) +
  theme_minimal()
```

```{r}
# Exploratory summaries from recent measles data

# Exploratory plots
# Plot 1
recent_state |>
  arrange(desc(total_cases_2024_26)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, total_cases_2024_26),
             y = total_cases_2024_26)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top states by measles cases (2024–2026)",
    x = "State",
    y = "Total cases"
  ) +
  theme_minimal()
```

```{r}
# Plot 2
##Select certain states to hone in on
focus_states <- c("Texas","South Carolina","Arizona","Utah","California","New York")

#Plot the measles trends for the six selected states, some of which are having active measles outbreaks in the U.S.
cdcmeaslesbystate_long |>
  filter(state %in% focus_states) |>
  ggplot(aes(year, cases, color = state)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(min(cdcmeaslesbystate_long$year),
                 max(cdcmeaslesbystate_long$year),
                 by = 1)
  ) +
  labs(
    title = "Year-by-year recent trends for key states",
    x = "Year",
    y = "Cases"
  ) +
  theme_minimal()
```
```{r}
# Exploratory summaries from historic measles data

# Plot 1
hist_state |>
  arrange(desc(pct_years_any)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, pct_years_any),
             y = pct_years_any)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "States with most persistent measles presence (1980–2009)",
    x = "State",
    y = "Percent of years with cases"
  ) +
  theme_minimal()

# Plot 2
hist_state |>
  arrange(desc(longest_run_nonzero)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(state, longest_run_nonzero),
             y = longest_run_nonzero)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Longest consecutive measles streak by state",
    x = "State",
    y = "Years in a row with measles"
  ) +
  theme_minimal()
```

```{r}
# Exploratory summaries from educational attainment dataset

# Plot 1: 18 to 24 year old percent graduation from high school by state in 2012
education_clean |>
  arrange(desc(`2012_18to24hs_percent`)) |>
  ggplot(aes(x = `2012_18to24hs_percent`,
             y = reorder(state, `2012_18to24hs_percent`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Top states by high school graduation rates in 2012",
    x = "State",
    y = "Percent"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Plot 2: 18 to 24 year old percent graduation from high school by state in 2022
education_clean |>
  arrange(desc(`2022_18to24hs_percent`)) |>
  ggplot(aes(x = `2022_18to24hs_percent`,
             y = reorder(state, `2022_18to24hs_percent`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Top states by high school graduation rates in 2022",
    x = "State",
    y = "Percent"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#Plot 3: High school graduation rate differences over time
education_clean |>
  arrange(desc(`hsgradratediff`)) |>
  ggplot(aes(x = `hsgradratediff`,
             y = reorder(state, `hsgradratediff`))) +
  geom_point() +
  coord_flip() +
  labs(
    title = "Graduation Rate Differences from 2012 to 2022",
    x = "State",
    y = "Percent Difference"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Exploratory data analysis

```{r}
# Recent measles cases vs average MMR coverage
# plot without linear regression
library(ggrepel)

features |>
  ggplot(aes(x = mmr_mean, y = total_cases_2024_26)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  scale_y_sqrt() +
  labs(
    title = "Recent measles cases vs average MMR coverage",
    x = "Average MMR coverage (%)",
    y = "Total measles cases 2024–2026 (sqrt scale)"
  ) +
  theme_minimal()
```

```{r}
# Recent measles cases vs average MMR coverage
# plot with linear regression
features |>
  ggplot(aes(x = mmr_mean, y = total_cases_2024_26)) +
  geom_point(size = 2) +
  
  # Add regression line with confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # Label each state without overlapping
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scale helps when some states have very high counts
  scale_y_sqrt() +
  
  labs(
    title = "Recent measles cases vs average MMR coverage",
    x = "Average MMR coverage (%)",
    y = "Total measles cases 2024–2026 (sqrt scale)"
  ) +
  theme_minimal()
```


```{r}
# Recent peak measles cases vs historical peak measles burden
# plot with linear regression
features |>
  ggplot(aes(x = max_cases_80_09, y = peak_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent peak measles cases vs historical peak measles burden",
    x = "Maximum yearly measles cases (1980–2009)",
    y = "Peak measles cases (2024–2026)"
  ) +
  
  theme_minimal()
```

```{r}
# MMR Vaccination Rates vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25pluslessthanhs_percent`, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "MMR Vaccine Rates correlated to 2022 rates of high school incompletion by state",
    x = "Incomplete High School Graduation Rates (%)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# Vaccination rates vs positive/negative difference in high school graduation rate by state
# plot with linear regression
features |>
  ggplot(aes(x = hsgradratediff, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent MMR vaccine rates correlated to differences in graduation rate between 2012 and 2022",
    x = "High School Graduation Rate Difference (%)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# MMR vaccination rate vs grad school completion
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25plusbachorhighertotal_percent`, y = mmr_mean)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "MMR Vaccination Rates correlated to attaining higher education degree(s)",
    x = "Bacherlors or Higher Education Rate (percent)",
    y = "Average MMR Coverage (%)"
  ) +

  theme_minimal()

# Peak measles case burden (2024-26) vs grad school completion
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25plusbachorhighertotal_percent`, y = peak_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Recent peak measles cases correlated to differences in graduation rate between 2012 and 2022",
    x = "High School Graduation Rate Difference (percent)",
    y = "Peak measles cases (2024–2026)"
  ) +

  theme_minimal()

# Total Measles Cases vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2022_25pluslessthanhs_percent`, y = total_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Total Measles Cases (2024-26) correlated to 2022 rates of high school incompletion by state",
    x = "Incomplete High School Graduation Rates (%)",
    y = "Total Measles Cases (2024-26)"
  ) +

  theme_minimal()

# Total Measles Cases vs Rate of Incomplete High School Graduation Rates
# plot with linear regression
features |>
  ggplot(aes(x = `2012_18to24hs_percent`, y = total_cases_2024_26)) +
  
  # scatter points
  geom_point(size = 2) +
  
  # regression line + confidence band
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  
  # state labels that avoid overlap
  geom_text_repel(
    aes(label = state),
    size = 3,
    max.overlaps = 50
  ) +
  
  # sqrt scales help when some states have extreme outbreak values
  scale_x_sqrt() +
  scale_y_sqrt() +
  
  labs(
    title = "Total Measles Cases (2024-26) correlated to 2022 rates of high school incompletion by state",
    x = "2012 High School Graduation Rates for 18-24 year olds (%)",
    y = "Total Measles Cases (2024-26)"
  ) +

  theme_minimal()
```

# **Data Analysis** 

## Regression analyses

# The following analysis examine bivariate relationships between recent measles burden (`total_cases_2024_26`) and selected predictors. For each model, we show a scatterplot with a fitted linear trend, a coefficient table from `tidy()`, and model fit statistics from `glance()`. 
```{r}
library(tidymodels)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(readr)

# Create analysis dataset with variables used in models
bivar_analysis_data <- features %>%
  select(
    state,
    total_cases_2024_26,
    mmr_mean,
    frac_years_below_95,
    hsgradratediff,
    pct_years_any,
    mean_cases_80_09,
    longest_run_nonzero
  ) %>%
  drop_na()
# Define linear regression model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")
```
# Bivariate linear regression models
```{r}
# Model 1: Total measles cases vs mean MMR coverage
# Fit bivariate model with mean MMR coverage as predictor

mmrbivar_fit <- lm_spec %>%
  fit(total_cases_2024_26 ~ mmr_mean, data = bivar_analysis_data )

# Plot
ggplot(bivar_analysis_data , aes(x = mmr_mean, y = total_cases_2024_26)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Mean MMR Coverage",
    x = "Mean MMR coverage",
    y = "Total measles cases, 2024-2026"
  ) +
  theme_minimal()

# Coefficient table
tidy(mmrbivar_fit) %>%
  kable(caption = "Tidy results: total_cases_2024_26 ~ mmr_mean")

# Model fit summary
glance(mmrbivar_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: total_cases_2024_26 ~ mmr_mean")

```
```{r}
# Model 2: Total measles cases vs fraction of years below 95% MMR

# Fit bivariate model with fraction of years below 95% as predictor
below95bivar_fit <- lm_spec %>%
  fit(total_cases_2024_26 ~ frac_years_below_95, data = bivar_analysis_data)

# Plot
plot_totalcases_below95 <- ggplot(bivar_analysis_data , aes(x = frac_years_below_95, y = total_cases_2024_26)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Fraction of Years Below 95% MMR",
    x = "Fraction of years below 95%",
    y = "Total measles cases, 2024-2026"
  ) +
  theme_minimal()
# Coefficient table
tidy(below95bivar_fit) %>%
  kable(caption = "Tidy results: total_cases_2024_26 ~ frac_years_below_95")

# Model fit summary
glance(below95bivar_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: total_cases_2024_26 ~ frac_years_below_95")
```

```{r}
# Model 3: Total measles cases vs historical measles persistence

# Fit bivariate model with historical measles persistence as predictor
historybivar_fit <- lm_spec %>%
  fit(total_cases_2024_26 ~ pct_years_any, data = bivar_analysis_data )

# Plot
ggplot(bivar_analysis_data , aes(x = pct_years_any, y = total_cases_2024_26)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Historical Measles Persistence",
    x = "Percent of historical years with any measles cases",
    y = "Total measles cases, 2024-2026"
  ) +
  theme_minimal()
# Coefficient table
tidy(historybivar_fit) %>%
  kable(caption = "Tidy results: total_cases_2024_26 ~ pct_years_any")
# Model fit summary
glance(historybivar_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: total_cases_2024_26 ~ pct_years_any")
```

```{r}
# Model 4: Total measles cases vs change in high school graduation rate

# Fit bivariate model with education change as predictor
edubivar_fit <- lm_spec %>%
  fit(total_cases_2024_26 ~ hsgradratediff, data = bivar_analysis_data)

# Plot
plot_totalcases_edu <- ggplot(bivar_analysis_data , aes(x = hsgradratediff, y = total_cases_2024_26)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Change in High School Graduation Rate",
    x = "High school graduation rate difference",
    y = "Total measles cases, 2024-2026"
  ) +
  theme_minimal()
plot_totalcases_edu
# Save plot
ggsave(here("results","figures","totalcases vs edu biovar.png"), plot = plot_totalcases_edu, width = 6, height = 4)

# Coefficient table
tidy(edubivar_fit) %>%
  kable(caption = "Tidy results: total_cases_2024_26 ~ hsgradratediff")

# Save results
write_csv(tidy(edubivar_fit), here("results", "tables", "tidy_totalcases vs edu biovar.csv"))

# Model fit summary
glance(edubivar_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: total_cases_2024_26 ~ hsgradratediff")
# Save results
write_csv(glance(edubivar_fit), here("results","tables","glance_totalcases vs edu biovar.csv"))
```
# Also wanted to see what the models and plots look like if they were log transformed (see below). Since the above models do not make intuitive sense. 

```{r}
log_bivar_analysis_data <- features %>%
  mutate(
    log_total_cases = log1p(total_cases_2024_26)
  )
```

```{r}
# Model 1: Total measles cases (log transformed) vs mean MMR coverage
# Fit bivariate model with mean MMR coverage as predictor

plot_totalcases_mmr <- mmrbivar_log_fit <- lm_spec %>%
  fit(log_total_cases ~ mmr_mean, data = log_bivar_analysis_data)

# Plot
plot_totalcases_mmr <-ggplot(log_bivar_analysis_data , aes(x = mmr_mean, y = log_total_cases)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Mean MMR Coverage",
    x = "Mean MMR coverage",
    y = "Total measles cases, 2024-2026 (log transformed)"
  ) +
  theme_minimal()
plot_totalcases_mmr
# Save plot
ggsave(here("results","figures","totalcases vs MMR biovar.png"), plot = plot_totalcases_mmr, width = 6, height = 4)

# Coefficient table
tidy(mmrbivar_log_fit) %>%
  kable(caption = "Tidy results: log_total_cases ~ mmr_mean")
# Save results
write_csv(tidy(mmrbivar_log_fit), here("results","tables","tidy_totalcases vs MMR biovar.csv"))

# Model fit summary
glance(mmrbivar_log_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: log_total_cases ~ mmr_mean")
# Save results
write_csv(glance(mmrbivar_log_fit), here("results","tables","glance_totalcases vs MMR biovar.csv"))
```
Biovariable Model 1: After log-transforming the data to handle outliers, this model shows that states with higher MMR vaccination rates generally tend to have fewer measles cases. However, this connection isn't strong enough to be definitive, as vaccination rates alone only explain about 4% of the differences seen between states.

```{r}
# Model 2: Total measles cases vs fraction of years below 95% MMR

# Fit bivariate model with fraction of years below 95% as predictor
below95bivar_log_fit <- lm_spec %>%
  fit(log_total_cases ~ frac_years_below_95, data = log_bivar_analysis_data)

# Plot
plot_totalcases_below95 <-ggplot(log_bivar_analysis_data , aes(x = frac_years_below_95, y = log_total_cases)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Fraction of Years Below 95% MMR",
    x = "Fraction of years below 95%",
    y = "Total measles cases, 2024-2026 (log transformed)"
  ) +
  theme_minimal()
plot_totalcases_below95
# Save plot
ggsave(here("results","figures","totalcases vs below95 biovar.png"), plot = plot_totalcases_below95, width = 6, height = 4)

# Coefficient table
tidy(below95bivar_log_fit) %>%
  kable(caption = "Tidy results: log_total_cases ~ frac_years_below_95")

# Save results
write_csv(tidy(below95bivar_log_fit), here("results","tables","tidy_totalcases vs below95 biovar.csv"))

# Model fit summary
glance(below95bivar_log_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: log_total_cases ~ frac_years_below_95")
# Save results
write_csv(glance(below95bivar_log_fit), here("results","tables","glance_totalcases vs below95 biovar.csv"))
```
Biovariable Model 2: This model shows that for every year a state's vaccination rate stays below the 95% target, there is a 242% increase in the number of predicted measles cases. While this is the strongest predictor found so far, these vaccination gaps still only explain about 7% of the total differences seen between states.
```{r}
# Model 3: Total measles cases (log transformed) vs historical measles persistence

# Fit bivariate model with historical measles persistence as predictor
historybivar_log_fit <- lm_spec %>%
  fit(log_total_cases ~ pct_years_any, data = log_bivar_analysis_data )

# Plot
plot_totalcases_history <-ggplot(log_bivar_analysis_data , aes(x = pct_years_any, y = log_total_cases)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Historical Measles Persistence",
    x = "Percent of historical years with any measles cases",
    y = "Total measles cases, 2024-2026 (log transformed)"
  ) +
  theme_minimal()
plot_totalcases_history
# Save plot
ggsave(here("results","figures","totalcases vs history biovar.png"), plot = plot_totalcases_history, width = 6, height = 4)

# Coefficient table
tidy(historybivar_log_fit) %>%
  kable(caption = "Tidy results: log_total_cases ~ pct_years_any")
# Save results
write_csv(tidy(historybivar_log_fit), here("results","tables","tidy_totalcases vs history biovar.csv"))

# Model fit summary
glance(historybivar_log_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: log_total_cases ~ pct_years_any")

# Save results
write_csv(glance(historybivar_log_fit), here("results","tables","glance_totalcases vs history biovar.csv"))

```
Biovariable Model 3:This model shows that states with a constant history of measles outbreaks can see a 656% increase in predicted total cases compared to states with no history of disease. Even though this history of disease presence is a better predictor than some other factors, it still only explains about 4.4% of why case numbers differ from state to state.```{r}
```{r}
# Model 4: Total measles cases (log transformed) vs change in high school graduation rate

# Fit bivariate model with education change as predictor
edubivar_log_fit <- lm_spec %>%
  fit(log_total_cases ~ hsgradratediff, data = log_bivar_analysis_data)

# Plot
plot_totalcases_log_edu <- ggplot(log_bivar_analysis_data , aes(x = hsgradratediff, y = log_total_cases)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Total Measles Cases vs Change in High School Graduation Rate",
    x = "High school graduation rate difference",
    y = "Total measles cases, 2024-2026"
  ) +
  theme_minimal()
plot_totalcases_log_edu
# Save plot
ggsave(here("results","figures","totalcases vs edu log trans biovar.png"), plot = plot_totalcases_log_edu, width = 6, height = 4)

# Coefficient table
tidy(edubivar_log_fit) %>%
  kable(caption = "Tidy results: log_total_cases ~ hsgradratediff")

# Save results
write_csv(tidy(edubivar_log_fit), here("results","tables","tidy_totalcases vs edu log trans biovar.csv"))

# Model fit summary
glance(edubivar_log_fit) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC, nobs) %>%
  kable(caption = "Model fit summary: log_total_cases ~ hsgradratediff")

# Save results
write_csv(glance(edubivar_log_fit), here("results","tables","glance_totalcases vs edu log trans biovar.csv")) 
```

Biovariable Model 4: This model shows that for every percentage point increase in a state's high school graduation rate, there is a 13.9% increase in the number of predicted measles cases. However, this link is very weak and only explains about 3% of the differences in case numbers between states.Therefore there is little evidence to suggest that graduation rate trends are a primary driver of measles outbreaks. 

# Multivariable analysis

```{r}
# log transform variable to minimize outliers

# Helper function
# signed_log1p(x) behaves like log1p(x) for positive values, but it also works for negative values:
#   signed_log1p(x) = sign(x) * log(1 + abs(x))
signed_log1p <- function(x) {
  sign(x) * log1p(abs(x))
}

set.seed(123)

# 4-PREDICTOR MODEL DATA

# Outcome:
#   total_cases_2024_26

# Predictors:
#   pct_years_any         = historical measles
#   mmr_mean              = mean MMR
#   frac_years_below_95   = below 95% MMR vaccination
#   hsgradratediff        = incomplete high school rate

main_data <- features %>%
  select(
    total_cases_2024_26,
    pct_years_any,
    mmr_mean,
    frac_years_below_95,
    hsgradratediff
  ) %>%
  drop_na() %>%
  mutate(
    log_total_cases = signed_log1p(total_cases_2024_26),
    log_pct_years_any = signed_log1p(pct_years_any),
    log_mmr_mean = signed_log1p(mmr_mean),
    log_frac_years_below_95 = signed_log1p(frac_years_below_95),
    log_hsgradratediff = signed_log1p(hsgradratediff)
  )

# Train/test split for the main 4-predictor model

main_split <- initial_split(main_data, prop = 0.75)

train_data <- training(main_split)
test_data  <- testing(main_split)


# 10-fold CV on training data only
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 10)


# Linear regression specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

# MAIN LOG-TRANSFORMED 4-PREDICTOR MODEL
main_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_pct_years_any +
      log_mmr_mean +
      log_frac_years_below_95 +
      log_hsgradratediff
  )

# Fit on training data
main_fit <- fit(main_wf, data = train_data)

# Training MAE
# Note: this MAE is on the log-transformed outcome scale

train_preds <- predict(main_fit, new_data = train_data) %>%
  bind_cols(train_data)

train_mae <- train_preds %>%
  mae(truth = log_total_cases, estimate = .pred)

train_mae


# Cross-validated MAE
main_cv <- fit_resamples(
  main_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

main_cv_metrics <- collect_metrics(main_cv)
main_cv_metrics


# Test MAE
test_preds <- predict(main_fit, new_data = test_data) %>%
  bind_cols(test_data)

test_mae <- test_preds %>%
  mae(truth = log_total_cases, estimate = .pred)

test_mae


# Summary table for the main model

main_mae_summary <- tibble(
  dataset = c("Training", "Cross-validation", "Test"),
  MAE = c(
    train_mae$.estimate,
    main_cv_metrics$mean[main_cv_metrics$.metric == "mae"],
    test_mae$.estimate
  )
)

main_mae_summary
# summary table when using all 4 predictors 

# TEST WHICH OF THE 4 MAIN PREDICTORS ARE BEST

# Idea:
# Compare the full 4-predictor model to models that remove
# one predictor at a time. If MAE gets higher when a predictor is removed, that predictor is helping.

no_hist_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_mmr_mean +
      log_frac_years_below_95 +
      log_hsgradratediff
  )

no_mmrmean_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_pct_years_any +
      log_frac_years_below_95 +
      log_hsgradratediff
  )

no_below95_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_pct_years_any +
      log_mmr_mean +
      log_hsgradratediff
  )

no_hsgrad_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_pct_years_any +
      log_mmr_mean +
      log_frac_years_below_95
  )

# Cross-validation for each reduced model
no_hist_cv <- fit_resamples(
  no_hist_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

no_mmrmean_cv <- fit_resamples(
  no_mmrmean_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

no_below95_cv <- fit_resamples(
  no_below95_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

no_hsgrad_cv <- fit_resamples(
  no_hsgrad_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

# Put CV MAE values into one comparison table
predictor_importance_table <- tibble(
  model = c(
    "Full 4-predictor log model",
    "Without pct_years_any",
    "Without mmr_mean",
    "Without frac_years_below_95",
    "Without hsgradratediff"
  ),
  cv_mae = c(
    collect_metrics(main_cv)$mean[collect_metrics(main_cv)$.metric == "mae"],
    collect_metrics(no_hist_cv)$mean[collect_metrics(no_hist_cv)$.metric == "mae"],
    collect_metrics(no_mmrmean_cv)$mean[collect_metrics(no_mmrmean_cv)$.metric == "mae"],
    collect_metrics(no_below95_cv)$mean[collect_metrics(no_below95_cv)$.metric == "mae"],
    collect_metrics(no_hsgrad_cv)$mean[collect_metrics(no_hsgrad_cv)$.metric == "mae"]
  )
) %>%
  mutate(
    mae_increase_vs_full = cv_mae - cv_mae[model == "Full 4-predictor log model"]
  ) %>%
  arrange(cv_mae)

# table for which of the 4 predictors performs the best
predictor_importance_table
# Save Table
write_csv(
  predictor_importance_table,
  here("results","tables","which_of_the_4_predictors_performs_best.csv")
)

  
# plot for predictor importance
predictor_importance_plot <- predictor_importance_table %>%
  filter(model != "Full 4-predictor log model") %>%
  ggplot(aes(x = reorder(model, mae_increase_vs_full), y = mae_increase_vs_full)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Predictor importance from change in CV MAE",
    x = "Model with predictor removed",
    y = "Increase in MAE vs full model"
  ) +
  theme_minimal()

predictor_importance_plot

# Save plot
ggsave(
  here("results","figures","Predictor_Importance.png"),
  plot = predictor_importance_plot,
  width = 6,
  height = 4
)
```

```{r}

# Two-predictor model using best predictors
# mean MMR + below 95% vaccination

signed_log1p <- function(x) {
  sign(x) * log1p(abs(x))
}

two_pred_data <- features %>%
  select(
    total_cases_2024_26,
    mmr_mean,
    frac_years_below_95
  ) %>%
  drop_na() %>%
  mutate(
    log_total_cases = signed_log1p(total_cases_2024_26),
    log_mmr_mean = signed_log1p(mmr_mean),
    log_frac_years_below_95 = signed_log1p(frac_years_below_95)
  )


# Train / test split
set.seed(123)

two_split <- initial_split(two_pred_data, prop = 0.75)

train_data <- training(two_split)
test_data  <- testing(two_split)


# Model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

two_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_total_cases ~
      log_mmr_mean +
      log_frac_years_below_95
  )


# Fit model

two_fit <- fit(two_wf, data = train_data)


# Training MAE
train_preds <- predict(two_fit, train_data) %>%
  bind_cols(train_data)

train_mae <- train_preds %>%
  mae(truth = log_total_cases, estimate = .pred)

train_mae



# 10-fold cross validation
set.seed(123)

cv_folds <- vfold_cv(train_data, v = 10)

cv_results <- fit_resamples(
  two_wf,
  resamples = cv_folds,
  metrics = metric_set(mae)
)

collect_metrics(cv_results)



# Test MAE
test_preds <- predict(two_fit, test_data) %>%
  bind_cols(test_data)

test_mae <- test_preds %>%
  mae(truth = log_total_cases, estimate = .pred)

test_mae



# MAE summary table
mae_summary <- tibble(
  dataset = c("Training", "Cross-validation", "Test"),
  MAE = c(
    train_mae$.estimate,
    collect_metrics(cv_results)$mean,
    test_mae$.estimate
  )
)

mae_summary
# Summary of how the performance of the two prediction model
# Save summary
write_csv(
  mae_summary,(here("results","tables","Two-Predictor_Model_Summary.csv")))

# Save summary
write_csv(
  mae_summary,
  here("results", "tables", "Two-Predictor_Model_Summary.csv")
)

# Create plot
two_predictor_actual_plot <- ggplot(train_preds, aes(x = log_total_cases, y = .pred)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Two-Predictor Model: Predicted vs Actual Measles Cases",
    subtitle = "Predictors: Mean MMR and Below-95% Vaccination Rate",
    x = "Actual (log-transformed measles cases)",
    y = "Predicted (log-transformed measles cases)"
  ) +
  theme_minimal()

# Show plot
two_predictor_actual_plot

# Save plot
ggsave(
  filename = here("results", "figures", "Two-Predictor_Model.png"),
  plot = two_predictor_actual_plot,
  width = 6,
  height = 4
)
```
The predicted versus actual plot shows points scattered around the 45-degree line, indicating that the model captures the general relationship between vaccination rates and measles cases. While predictions are not exact, the diagonal trend suggests the two predictors provide meaningful predictive information.